Examining background gene lists in clusterProfiler - RNA-seq example

### 2024-06-05 analysis/analyse_rnaseq1.Rmd
improve charts

names(gs) <- gsub(".v2023.2.Hs.symbols.gmt","",names(gs))

@@

# bad=ORA analysis, a list with results of up and down regulated pathways
# good=same as bad, except fixed the background problem
compare_ora_plots <- function(bad,good){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  mup <- merge(badup,goodup,by=0)
  MAX=max(c(mup$es.x,mup$es.y))

  plot(mup$es.x, mup$es.y,
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="Enrichment score")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

@@

  plot(-log10(mup$pvalue.x), -log10(mup$pvalue.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="-log10 p-value")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

@@

  plot(mdn$es.x, mdn$es.y,
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="Enrichment score")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

@@

  plot(-log10(mdn$pvalue.x), -log10(mdn$pvalue.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original", ylab="corrected", main="-log10 p-value")
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)
}

@@

par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
barplot(t(df),beside=TRUE,horiz=TRUE,las=1,
  xlab="no. significant sets", main="Comparison of original and corrected data",
  col=c("gray30","gray80"))

legend("bottomright", inset=.02, legend=c("original","corrected"),
  fill=c("gray30","gray80"), horiz=TRUE, cex=1)

# RATIO
rat <- matrix( (res1$nsig_good+0) / (res1$nsig_bad+0) , ncol=9 )
rownames(rat) <- names(d)
colnames(rat) <- names(gs)

signif(rat,3) %>%
  kbl(row.names = TRUE, caption="Ratio of significant genesets Corrected:Original") %>%
  kable_paper("hover", full_width = F)

rat2 <- apply(rat,2,function(x) { x <- x[!is.na(x)] ; x <- x[is.finite(x)] ;  mean(x) } )
rat2

barplot(sort(rat2),horiz=TRUE,las=1,main="increase in significant sets after correction",xlab="fold increase")

@@

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]])
  good <- oragood(d[[1]],gs[[i]])
  compare_ora_plots(bad,good)
} )

>>>>

names(gs)
names(gs) <- c("KEGG", "Reactome", "Wikipathways", "miR targets", "TFT GTRD",
  "GO", "HPO", "Cellmarkers", "Hallmark" )

@@

# bad=ORA analysis, a list with results of up and down regulated pathways
# good=same as bad, except fixed the background problem
compare_ora_plots <- function(bad,good,libname=NULL){
  badup <- bad$orabad_up
  baddn <- bad$orabad_dn
  goodup <- good$oragood_up
  gooddn <- good$oragood_dn
  mup <- merge(badup,goodup,by=0)
  MAX=max(c(mup$es.x,mup$es.y))
  UPHEADER=paste(libname,"up")
  DNHEADER=paste(libname,"dn")

  par(mfrow=c(2,2))

  plot(mup$es.x, mup$es.y,
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original ES", ylab="corrected ES", main=UPHEADER)
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

@@

  plot(-log10(mup$pvalue.x), -log10(mup$pvalue.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original -log10 p-value", ylab="corrected -log10 p-value", main=UPHEADER)
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

@@

  plot(mdn$es.x, mdn$es.y,
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original ES", ylab="corrected ES", main=DNHEADER)
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)

@@

  plot(-log10(mdn$pvalue.x), -log10(mdn$pvalue.y),
    xlim=c(0,MAX), ylim=c(0,MAX),
    xlab="original -log10 p-value", ylab="corrected -log10 p-value", main=DNHEADER)
  grid()
  abline(a = 0, b = 1,lwd=2,lty=2)
}

@@

par(mar=c(c(5.1, 7.1, 4.1, 2.1) ))
barplot(t(df),beside=TRUE,horiz=TRUE,las=1,xlim=c(0,1250),
  xlab="no. significant sets", main="Effect of correcting the background bug",
  col=c("gray30","gray80"))
text(df$ns1+50,((1:nrow(df))*3)-1.5,labels=signif(df[,1],3))
text(df$ns2+50,((1:nrow(df))*3)-0.5,labels=signif(df[,2],3))
#abline(v=seq(100,1000,100),lty=2,lwd=0.5,col="gray")
legend("bottomright", inset=.02, legend=c("corrected","original"),
  fill=c("gray80","gray30"), horiz=FALSE, cex=1.1)

mylabels <- gsub("$","%",gsub("^","+",signif(((df$ns2/df$ns1)-1)*100,3)))
text(df$ns2+160,((1:nrow(df))*3)-1,labels=mylabels,cex=1.1)

@@

lapply(1:length(gs), function(i) {
  bad <- orabad(d[[1]],gs[[i]])
  good <- oragood(d[[1]],gs[[i]])
  compare_ora_plots(bad,good,libname=names(gs)[i])
} )
